# Birthday paradox

The birthday problem is a classic probability puzzle that demonstrates
how likely it is for at least two people in a group to share the same
birthday. While it might seem intuitive that the probability is low in
small groups, the results are surprising.

::: {#exm-birthday}
In a group of $k$ people, what is the probability that at least two
people share the same birthday? Assume (1) there are 365 possible
birthdays; (2) birthdays are evenly distributed across the year; (3)
people are equally likely to be born on any given day.
:::

::: solution
If $k>365$, the probability is $1$. Assume $k\leq365$ for the rest.
Instead of directly calculating the probability of at least two people
sharing a birthday, we first compute the complementary probability,
$P(\text{no match})$ , where no two people in the group have the same
birthday.

For the first person, there are 365 choices for their birthday. For the
second person, to avoid a shared birthday, there are 364 remaining
choices. For the third person, there are 363 choices, and so on. For $k$
people, the total number of arrangements (no shared birthday) is:

$$365 \times 364 \times 363 \times \cdots \times (365 - k + 1)$$

The total number of possible arrangements (with or without shared
birthdays) is $365^k$. Thus, the probability of no shared birthday is:

$$P(\textrm{no match})=\frac{365\cdot364\cdots(365-k+1)}{365^{k}}$$

Thus, the probability of at least one matched birthday is:
$$P(\text{match})= 1- P(\text{no match})=\begin{cases}
50.7\% & k=23\\
70.6\% & k=30\\
97.0\% & k=50\\
99.9\% & k=70
\end{cases}$$
:::

## R simulation {.unnumbered}

```{r}
# a class of k people 
k <- 30

# number of experiments
N <- 1000

# number of matches
m <- 0

for (i in 1:N) {
  
  # draw k random numbers from 1 to 365 with replacement
  birthdays <- sample(1:365, k, replace = T) 
    
  # number of duplicated birthdays
  dups <- duplicated(birthdays)
    
  # if duplicated birthdays found
  if (any(dups)) m <- m + 1
  
}

cat("Prob of at least one match: ", m / N)

```

There is even a built-in function `pbirthday` that computes the
probability of birthday coincidence. We may utilize this function to
plot the probability as the number of people increases.

```{r, results='hide'}
# compute the probability of birthday match for 30 people
prob <- pbirthday(30) 

# compute a vector of probabilities for 1,2...100 people
probs <- sapply(1:100, pbirthday)

# make a plot
plot(1:100, probs, main = "Probability of >1 people with the same birthday")
```
