# Data descriptives*

A random variable is a mathematical abstraction that provides a bridge
between theoretical probability and real-world data. Every dataset can 
be viewed as observations from random variables.

Despite the outcome of any one event being uncertain, we can use
patterns from past observations to predict the general behavior of these
variables. By collecting data, we can figure out how often certain
outcomes occur and connect them to theoretical distributions.

$$\begin{array}{ccccccc}
\textrm{Question with} & \rightarrow & \textrm{Data} & \rightarrow & \textrm{Patterns}\\
\textrm{uncertainty} &  & \downarrow &  & \downarrow\\
 &  & \textrm{RVs} & \rightarrow & \textrm{Distributions} & \rightarrow & \textrm{Predictions}\\
\end{array}$$

**Columns as random variables**

In a dataset, we view every column as a random variable.

```{r}
# Load a dataset from a CSV file
exam <- read.csv("../dataset/exam.csv")

# View the data: each column is a random variable
head(exam)
```

**Summary statistics**

We can describe the distribution of a variable with summary statistics: 
such as quartiles, deciles and percentiles.

```{r}
summary(exam$overall)
```


**Histograms**

One way to visualize the distribution of a variable is to plot a
histogram. A histogram groups data points into intervals, showing how
often data values fall within each range. The horizontal axis represents
the intervals (or bins), and the vertical axis shows the frequency or
count of data points in each bin.

A histogram gives an approximation of the true (unknown) distribution.
It is not itself the distribution. A distribution refers to the
*theoretical* assignment of probabilities across all possible outcomes,
whereas a histogram represents the *empirical* frequencies observed in a
finite sample.

```{r}
library(lattice)

histogram(~final, exam)
```

To explore conditional probability, we may plot the histogram
conditioned on other variables.

```{r}
histogram(~final | gender, exam)
```

**Boxplots**

A boxplot, also known as a box-and-whisker plot, displays the
median, quartiles, and range of the data. The box represents the
interquartile range (IQR), which contains the middle 50% of the data,
with the lower and upper edges corresponding to the first (Q1) and third
quartiles (Q3). Whiskers extend from the box to indicate the range of
values within 1.5 times the IQR from Q1 and Q3, while points beyond this
range are considered outliers.

```{r}
bwplot(final ~ major, exam)
```

**Scatter plots**

To observe the relationship between variables, it is straightforward to
make a scatter plot of $Y$ against $X$. An upward-sloping pattern in the
scatter plot indicates that the variables tend to move together, whereas
a downward-sloping pattern suggests that they move in opposite
directions. A flat slope implies the absence of correlation between the
two variables.

```{r}
xyplot(final ~ mid, exam)
```
