# Dice rolling formula\*

The Binomial distribution gives the formula for the probability of
observing $k$ heads when flipping $n$ coins. Can we find a formula for
the probability of getting a total of $p$ points when rolling $n$ dice?

The probability of obtaining $p$ points on $n$ $s$-sided dice can be
computed as the coefficient of $x^p$ in

$$f(x)=(x+x^2+...+x^s)^n$$

since each possible arrangement contributes one term.

$$f(x) = x^n(1+x+\dots+x^{s-1})^n = x^n\left(\frac{1-x^s}{1-x}\right)^n$$

To obtain the coefficient of $x^p$, expand the binomial power:

$$x^n(1-x^s)^n(1-x)^{-n} = x^n \sum_{k=0}^n (-1)^k \binom{n}{k} x^{sk}
\sum_{l=0}^\infty\binom{n+l-1}{l}x^l$$

The coefficient of $x^p$ include all terms with $p=n+sk+l$. Therefore,

$$c_p = \sum_{k=0}^n (-1)^k \binom{n}{k}\binom{p-sk-1}{p-sk-n}$$

But $p-sk-n>0$ only when $k<(p-n)/s$, so the other terms do not
contribute. Furthermore, applying the symmetric property of the binomial
formula, we have

$$\binom{p-sk-1}{p-sk-n}= \binom{p-sk-1}{n-1}$$

Therefore, the probability of getting $p$ points when rolling $n$
$s$-sided dice is given by

$$f(p,n,s) = \sum_{k=0}^{\lfloor{(p-n)/s}\rfloor} (-1)^k \binom{n}{k} \binom{p-sk-1}{n-1}.$$

::: callout-note
## Binomial formula for negative $n$

$$\begin{aligned}
\binom{-n}{k} &= \prod_{i=0}^{k-1} \frac{-n-i}{k-i}
=(-1)^k \prod_{i=0}^{k-1} \frac{n+i}{k-i} \\
&=(-1)^k\frac{n(n+1)\dots(n+k-1)}{k!}\\
&=(-1)^k\frac{(n+k-1)!}{k!(n-1)!}\\
&=(-1)^k\binom{n+k-1}{k}
\end{aligned}$$
:::

We can verify our formula by simulating the dice rolling game.

```{r}
set.seed(0)

# simulates rolling n dice and returns the sum
roll_dice <- function(n, s=6) {
  sum(sample(seq(1,s), n, replace = T))
}

# rolling 10 dice 1000 times and collect the results
points <- replicate(1000, roll_dice(10))

# distribution of the sum of points
hist(points, freq = F)
```

```{r}
# formula for computing probability of dice points
dice_formula <- function(p, n, s=6) {
  prob <- 1/s^n*sum(
    sapply(seq(0, floor((p-n)/s)), 
           function(k) (-1)^k*choose(n,k)*choose(p-s*k-1,n-1)))
}

# computing the probability of getting 20-50 when rolling 10 dice
x <- 20:50; 
y <- sapply(x, function(p) dice_formula(p,n=10))

# overlay the formula on the histogram
# it turns out the formula does a nice job!
hist(points, ylim = c(0, 0.07), freq = F)
lines(x, y, col = 2, lwd=2)
```
