# Earthquake simulation

::: {#exm-4.21}
What is the probability of $k$ earthquakes occurred in a random year in
Sichuan?

This is the case of rare events with approximately constant rate of
occurrence. We intend to model this with the Poisson distribution. The
key is to figure out the parameter $\lambda$.

Historical records[^25_earthquake-1] show that, from 26 BCE to 2021 CE,
there were 309 earthquakes with magnitude of 5.0 or greater. Let $X$ be
the number of earthquakes with magnitude 5.0 or greater. The annual rate
$\lambda$ of earthquakes is therefore $\frac{309}{2048}=0.15$.

Assume earthquakes are independent events (not always the case). Then
$X\sim\textrm{Pois}(0.15)$. By the distribution of the Poisson
distribution, $$P(X=k)=\begin{cases}
0.86 & k=0\\
0.13 & k=1\\
0.01 & k=2
\end{cases}.$$
:::

[^25_earthquake-1]: See [this
    article](https://www.scdzj.gov.cn/dzpd/dzkp/202201/t20220125_51277.html)
    from the Sichuan Earthquake Administration.

This is of course a rough approximation. Let's see how close the
occurrences of earthquakes simulated by a Poisson process to the actual
patterns of events.

```{r}
library(lattice)
library(data.table)

# Earthquakes of Sichuan 1900-2019. source: USGS
eq <- fread("../dataset/earthq.csv")

# only consider strong earthquakes with magnitude >= 6
# count the number of earthquakes by year
eq6 <- eq[mag >=6, .N, by = year] 
  
# plot the occurrence of earthquakes by year
xyplot(N~year, data=eq6, type="h", ylim=c(0,4))
```

```{r}
# earthquakes are likely to be followed by earthquakes
# the rate of occurrence, however, looks constant 
t <- seq(1900, 2019)
rate <- 0.3

# initialize the random generator
set.seed(3000)

# model the occurrence with Poisson distribution 
y <- rpois(length(t), rate)

# plot the simulated occurrence of earthquakes
xyplot(y ~ t, type="h", ylim = c(0,4))
```

The simulated result looks somewhat similar to the history. However,
there are a few notable difference. The simulated ones are distributed
more evenly, whereas in reality, earthquakes are not independent.
Moreover, the Poisson process seems to generate less rare events (more
than 4 earthquakes a year) than the reality. 

```{r, include=FALSE}
# generate a new sequence
z <- numeric(length(t))
z[1] <- rpois(1, rate)

for (i in 2:length(t)) {
  # introduce weak dependency
  # if an event occurs more often than expected, increase the 
  # rate of the next event; decrease the rate otherwise. 
  z[i] <- rpois(1, rate + 0.1*(z[i-1]-rate))  
}

# plot the series
xyplot(z ~ t, type = "h")
```
