## Covariance

For more than one random variable, it is also of interest to know the
relationship between them. Are they dependent? How strong is the
dependence? Covariance and correlation are intended to measure that
dependence. But they only capture a particular type of dependence,
namely linear dependence.

::: {#def-cov}
### Covariance
The covariance between random variables $X$ and $Y$ is defined as
$$Cov(X,Y)=E[(X-EX)(Y-EY)].$$
:::

The covariance between $X$ and $Y$ reflects how much $X$ and $Y$
*simultaneously* deviate from their respective means. 

::: {#thm-cov}
For any random variables $X$ and $Y$, $$Cov(X,Y)=E(XY)-E(X)E(Y).$$
:::

::: proof
Let $\mu_{X}=E(X)$ and $\mu_{Y}=E(Y)$. By definition, $$\begin{aligned}
Cov(X,Y) & =E(XY-\mu_{X}Y-\mu_{Y}X+\mu_{X}\mu_{Y})\\
 & =E(XY)-\mu_{X}E(Y)-\mu_{Y}E(X)+\mu_{X}\mu_{Y}\\
 & =E(XY)-E(X)E(Y).\end{aligned}$$
:::

::: {#thm-indep-cov}
If $X,Y$ are independent, they are uncorrelated. But the converse is
false.
:::

::: proof
1. $Cov(X,Y)=E(XY)-E(X)E(Y)$. Independence implies $E(XY)=E(X)E(Y)$. Thus,
$Cov(X,Y)=0$. 
2. $Cov(X,Y)=0$ does not necessarily imply independence.
Consider the following counter example. Let $X$ be a random variable
that takes three values -1, 0, 1 with equal probability. And $Y=X^{2}$.
$X$ and $Y$ are clearly dependent. But they their covariance is 0.
Since $E(X)=0$, $E(Y)=2/3$, $E(XY)=E(X^{3})=0$, $Cov(X,Y)=0$.
:::

::: callout-note
## Linear dependency
Covariance and correlation provide measures of the extend to which two
random variables are linearly related. It is possible that the
covariance is $0$ even when $X$ and $Y$ are dependent but the relationship is
nonlinear.

```{r, echo=FALSE, fig.asp=0.4}
# Set seed for reproducibility
set.seed(123)

# Number of observations
n <- 100

# Linear relationship
X1 <- rnorm(n)
Y1 <- 2 * X1 + rnorm(n, sd = 0.5)

# Nonlinear relationship
X2 <- rnorm(n)
Y2 <- X2^2 + rnorm(n, sd = 0.5)

# Independent variables
X3 <- rnorm(n)
Y3 <- rnorm(n)

# Visualize the relationship
# Set up a 1x3 grid for plots
par(mfrow = c(1, 3))
options(digits = 3)

# Plot 1: Linear Relationship
plot(X1, Y1, main = sprintf("Cov=%.3f",cov(X1,Y1)), pch = 16)
abline(lm(Y1 ~ X1), col = "red", lwd = 2)

# Plot 2: Nonlinear Relationship
plot(X2, Y2, main = sprintf("Cov=%.3f",cov(X2,Y2)), pch = 16)
curve(x^2, add = TRUE, col = "red", lwd = 2)

# Plot 3: Independent Variables
plot(X3, Y3, main = sprintf("Cov=%.3f",cov(X3,Y3)), pch = 16)
abline(h = mean(Y3), col = "red", lwd = 2)
```
:::

::: {#prp-cov}
Covariance has the following properties:

-   $Cov(X,X)=Var(X)$
-   $Cov(X,Y)=Cov(Y,X)$
-   $Cov(cX,Y)=Cov(X,cY)=c\left[Cov(X,Y)\right]$
-   $Cov(X+Y,Z)=Cov(X,Z)+Cov(Y,Z)$
-   $Var(X+Y)=Var(X)+Var(Y)+2Cov(X,Y)$

:::

::: proof
We only prove the variance-covariance property: $$\begin{aligned}
Var(X+Y) & =E[(X+Y-\mu_{X}-\mu_{Y})^{2}]\\
 & =E[(X-\mu_{X})^{2}+(Y-\mu_{Y})^{2}+2(X-\mu_{X})(Y-\mu_{Y})]\\
 & =Var(X)+Var(Y)+2Cov(X,Y).\end{aligned}$$
:::

::: {#thm-var-cov}
For random variables $X_1,X_2,\dots,X_n$, it holds that
$$Var\left(\sum_{i=1}^{n}X_{i}\right)=\sum_{i=1}^{n}Var(X_{i})+
2\sum_{i<j}Cov(X_{i},X_{j}).$$

If $X_1,X_2,\dots,X_n$ are identically distributed and have the same
covariance relationships (symmetric), then
$$Var\left(\sum_{i=1}^{n}X_{i}\right)=nVar(X_1)+2\binom{n}{2}Cov(X_1,X_2).$$
:::


While $\text{Cov}(X,Y)$ quantifies how $X$ and $Y$ vary together, its
magnitude also depends on the absolute scales of $X$ and $Y$ (multiply
$X$ by a constant $c$, the covariance will be different). To establish a
measure of association between $X$ and $Y$ that is unaffected by
arbitrary changes in the scales of either variable, we introduce a
"standardized covariance" called correlation.

::: {#def-corr}
### Correlation
The correlation between random variables $X$ and $Y$ is defined as
$$Corr(X,Y)=\frac{Cov(X,Y)}{\sqrt{Var(X)Var(Y)}}.$$
:::

By convention, we denote correlation by Greek letter $\rho\equiv Corr(X,Y)$.

Unlike covariance, scaling $X$ or $Y$ has no effect on the correlation.
We can verify this:
$$Corr(cX,Y)=\frac{Cov(cX,Y)}{\sqrt{Var(cX)Var(Y)}}=\frac{cCov(X,Y)}{c\sqrt{Var(X)Var(Y)}}=Corr(X,Y).$$

::: {#thm-corr}
For any random variable $X$ and $Y$, $$-1\leq Corr(X,Y)\leq1.$$
:::

::: proof
Without loss of generality, assume $X,Y$ both have variance 1, since
scaling does not change the correlation. Let $\rho=Corr(X,Y)=Cov(X,Y)$.
Then $$\begin{aligned}
Var(X+Y) & =Var(X)+Var(Y)+2Cov(X,Y)=2+2\rho\geq0,\\
Var(X-Y) & =Var(X)+Var(Y)-2Cov(X,Y)=2-2\rho\ge0.\end{aligned}$$ Thus
$-1\leq\rho\leq1$.
:::

-   $X$ and $Y$ are **positively correlated** if $\rho_{XY}>0$;
-   $X$ and $Y$ are **negatively correlated** if $\rho_{XY}<0$;
-   $X$ and $Y$ are **uncorrelated** if $\rho_{XY}=0$.

::: {#thm-corrlin}
Suppose that $X$ is a random variable and $Y=aX+b$ for some constants
$a,b$, where $a\neq0$. If $a>0$, then $\rho_{XY}=1$. If $a<0$, then
$\rho_{XY}=-1$.
:::

::: proof
If $Y=aX+b$, then $E(Y)=aE(X)+b$. Thus, $Y-E(Y)=a(X-E(X))$. Therefore,
$$Cov(X,Y)=aE[(X-EX)^{2}]=aVar(X).$$ Since $Var(Y)=a^{2}Var(X)$,
$\rho_{XY}=\frac{a}{|a|}$. The theorem thus follows.
:::

::: callout-note
### Correlation analysis
A **correlation matrix** shows the pairwise correlation coefficients between variables.
It’s one of the most common tools for exploring relationships in multivariate data.

```{r, message=FALSE}
# variables for analysis
vars <- mtcars[, 1:4]

# compute the correlation matrix
print(cor(vars))
```
:::

::: {#exm-4.16}
Let $X\sim \text{HGeom}(w,b,n)$. Find $Var(X)$.
:::

::: solution
Interpret $X$ as the number of white balls in a sample of size $n$ from
an box with $w$ white and $b$ black balls. We can represent $X$ as the
sum of indicator variables, $X=I_{1}+\cdots+I_{n}$ , where $I_{j}$ is
the indicator of the $j$-th ball in the sample being white. Each $I_{j}$
has mean $p=w/(w+b)$ and variance $p(1-p)$, but because the $I_{j}$ are
dependent, we cannot simply add their variances. Instead,
$$\begin{aligned}
Var(X) & =Var\left(\sum_{j=1}^{n}I_{j}\right)\\
 & =Var(I_{1})+\cdots+Var(I_{n})+2\sum_{i<j}Cov(I_{i},I_{j})\\
 & =np(1-p)+2\binom{n}{2}Cov(I_{i},I_{j})\end{aligned}$$

In the last step, because of symmetry, for every pair $i$ and $j$,
$Cov(I_{i},I_{j})$ are the same. $$\begin{aligned}
Cov(I_{i},I_{j}) & =E(I_{i}I_{j})-E(I_{i})E(I_{j})\\
 & =P(i\textrm{ and }j\textrm{ both white})-P(i\textrm{ is white})P(j\textrm{ is white})\\
 & =\frac{w}{w+b}\cdot\frac{w-1}{w+b-1}-p^{2}\\
 & =p\frac{Np-1}{N-1}-p^{2}\\
 & =\frac{p(p-1)}{N-1}\end{aligned}$$

where $N=w+b$. Plugging this into the above formula and simplifying, we
eventually obtain
$$Var(X)=np(1-p)+n(n-1)\frac{p(p-1)}{N-1}=\frac{N-n}{N-1}np(1-p).$$ This
differs from the Binomial variance of $np(1-p)$ by a factor of
$\frac{N-n}{N-1}$. This discrepancy arises because the Hypergeometric
story involves sampling without replacement. As $N\to\infty$, it becomes
extremely unlikely that we would draw the same ball more than once, so
sampling with or without replacement essentially become the same.
:::
