# Sum of random variables

We have seen the sum of coin heads and the sum of dice points follow a
"bell-shaped" distribution. This is not a coincidence. The pattern does
not exist when the number of coins or dice are small, but becomes
apparent when the numbers get large.

```{r}
# Function to simulate the sum of rolling n fair dice
simulate_dice_sum <- function(n, n_simulations) {
  
  # Simulate rolling n dice n_simulations times
  dice_rolls <- matrix(sample(1:6, n * n_simulations, replace = TRUE), 
                       nrow = n_simulations, ncol = n)
  
  # Compute the sum of each roll
  sums <- rowSums(dice_rolls)
  
  return(sums)
}

# Set seed for reproducibility
set.seed(123)

# Rolling 2 dice
sums <- simulate_dice_sum(2, 1000)

# Plot the distribution of sums
hist(sums, prob = TRUE, col = "lightblue")

# Rolling 10 dice
sums <- simulate_dice_sum(10, 1000)

# Plot the distribution of sums
hist(sums, prob = TRUE, ylim = c(0,.08), col = "lightblue")

# Overlay the normal curve
curve(dnorm(x, mean = mean(sums), sd = sd(sums)), 
      col = "red", lwd = 2, add = TRUE)
```

This does not only hold for dice points. In fact, the sum of random
variables from any distribution would reveal a similar pattern.

```{r}
# Sum of uniform random variables
simulate_uniform_sum <- function(n, n_simulations) {
  sums <- numeric(n_simulations)
  
  # Simulate the experiment n_simulations times
  for (i in 1:n_simulations) {
    # Sum of n Uniform random variables
    sums[i] <- sum(runif(n))
  }
  
  return(sums)
} 

# Sum of 2 Uniform random variables
sums <- simulate_uniform_sum(2, 1000)

# Plot the histogram
hist(sums, prob = TRUE, col = "lightblue")

# Sum of 10 Uniform random variables
sums <- simulate_uniform_sum(10, 1000)

# Plot the histogram
hist(sums, prob = TRUE, col = "lightblue")

# Overlay the normal curve
curve(dnorm(x, mean = mean(sums), sd = sd(sums)), 
      col = "red", lwd = 2, add = TRUE)
```

::: callout-note
## Sum of random variables approaches Normal distribution

Let $X_1, X_2, \dots, X_n$ be a sequence of **i.i.d** random variables
with mean $\mu=E(X_i)$ and variance $\sigma^2=Var(X_i)$. Let
$$S_n = X_1 + X_2 + \dots + X_n$$ Then, as $n\to\infty$, $S_n$
**converge in distribution** to a normal distribution. That is
$$S_n \to^d N(n\mu, n\sigma^2).$$
:::
