# Sum of RVs

We know the sum of coin heads and the sum of dice points follow a
Binomial distribution.

```{r}
set.seed(0)

# Repeat experiment: sum of heads tossing ten coins
S1 <- replicate(1000, { sum(sample(0:1, 20, TRUE)) })

hist(S1, prob=TRUE)

# Bell-shaped curve
curve(dnorm(x, mean(S1), sd(S1)), col=2, add=TRUE)
```

The distribution of the sum of dice has a similar shape:

```{r}
set.seed(0)

# Repeat experiment: sum of rolling ten dice
S2 <- replicate(10000, { sum(sample(1:6, 10, TRUE)) })

hist(S2, prob=TRUE)

# Bell-shaped curve
curve(dnorm(x, mean(S2), sd(S2)), col=2, add=TRUE)
```

This is not a coincidence. In fact, the sum of random variables 
from any distribution would reveal a similar shape.

```{r}
set.seed(0)

# Repeat experiment: sum of ten uniform variables
S3 <- replicate(1000, { sum(runif(10)) })

hist(S3, prob=TRUE)

# Bell-shaped curve
curve(dnorm(x, mean(S3), sd(S3)), col=2, add=TRUE)
```

The bell-shaped curve comes from Normal distributions. 
The distribution of the *sum* of random variables always converges to 
Normal distribution.

::: callout-note
## Sum of random variables approaches Normal

Let $X_1, X_2, \dots, X_n$ be a sequence of independent and identical random variables.
Let $$S_n = X_1 + X_2 + \dots + X_n$$ Then, as $n\to\infty$, $S_n$ converges to
Normal in distribution. That is
$$S_n \to^d \textrm{Normal}.$$
:::
